Two Decades of Smalltalk VM Development

### 2018-08-16 paper.tex
Rename to Two Decades Of Smalltralk Development Open-Smalltalk-VM => OpenSmalltalk-VM Add .gitignore; rm --cached the generated outputs, except the finished .pdf
Work on history section, snapshots and development.  Adding references for DeutschSchiffman, Spur, etc.

\def\openSmalltalkVM{Open-Smalltalk-VM\xspace}
%%%
% End Legend
%%%

%% Title information
\title[Over Twenty Years of Virtual Machine Development Through Simulation]{Over Twenty Years of Virtual Machine Development and Debugging Through Simulation}

@@

\openSmalltalkVM was originally written in Smalltalk to specify and example the virtual machine (VM) implementation. The same code base was changed a little bit to allow Smalltalk-to-C compilation to generate the production VM through the C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

As the VM evolved, by introducing better garbage collector algorithms or a just-in-time compiler, both execution models co-evolved. For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.

@@

To specify the language and explain how to write the virtual machine (VM) for it, the Smalltalk-80 crew wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample}. The VM was ported from the original specifications. Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture} and generate an efficient VM through the C compiler. Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM. At this point, the VM consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a generational garbage collector (GC) and
\item BitBlt, an extension for the user interface bit-based engine.
\end{itemize} 
These three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}. As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM. The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). The existing simulator was not able to simulate the machine code generated by the JIT and had to be extended by binding multiple processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).

@@

As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C. The Core VM code (the interpreter, the JIT and the memory manager), is written in Slang. The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C. 
%EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.

@@

There are multiple reasons why we keep writing the core components in Slang and not in C/C++. Most of them are small details. For instance, the Slang-to-C compiler generates C code slightly different from the Slang code using different annotations. It can for example duplicate the implementation of specific methods with specific constant operands to generate more efficient code in the interpreter. The main reason why we keep using Slang over C/C++ is VM simulation. By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing to develop and debug it with the Smalltalk development tools and a live programming experience.

@@

\paragraph{Start-up from snapshots.} 
Smalltalk is a snapshot-based language. A Smalltalk program is started from a snapshot, \emph{i.e.,} a memory dump of all live objects at a given point in the execution of a program. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory and resumes execution in the process active at snapshot time.

When programming with Smalltalk, the programmers usually starts from a snapshot which contains the core libraries, the development environment and the application developed. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the set of live objects. When the changes are done, the developer takes a new snapshot, which includes its changes. Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).

Starting up \openSmalltalkVM from source files has never been possible. However, recent work~\cite{PharoBootstrap} allowed one of the Smalltalk runtime to recreate a snapshot from sources, indirectly allowing to start the VM from source files.

\paragraph{Interpreter-only simulation.} 
As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.

>>>>

\def\openSmalltalkVM{OpenSmalltalk-VM\xspace}
%%%
% End Legend
%%%

%% Title information
\title[Two Decades of Smalltalk VM Development]{Two Decades of Smalltalk VM Development}

@@

\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator".  The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

As the VM has evolved, including adding a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved. For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

To specify the the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample} using the code from \cite{blueBook} as a starting point. Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.  Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process. Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM. At this point, the VM consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation, a pointer-reversing tracing garbage collector and a heap divided into two generations
\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine.
\item external C code to support running the VM on popular platforms
\end{itemize} 
The first three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk and Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM. One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). The code base was liberally sprinkled with assertions.  The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  A new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup. Bochs was used a second time for x64 support.

@@

As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C. The Core VM code (the interpreter, the JIT and the memory manager), is written in Slang. The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C, with a corresponding Smalltalk version for simulation. 
%EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.

@@

There are multiple reasons why we keep writing the core components in Slang and not in C/C++. Most of them are small details. For instance, the Slang-to-C compiler generates C code slightly different from the Slang code using different annotations. It can for example duplicate the implementation of specific methods with specific constant operands to generate more efficient code in the interpreter. The main reason why we keep using Slang over C/C++ is VM simulation. By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing development and debugging with the Smalltalk development tools and a live programming experience.

@@

\paragraph{Start-up from snapshots.} 
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory and resumes execution in the process active when the snapshot was created.  A new snapshot can be made during or at the end of the session.

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.  When the programmer chooses, the developer takes a new snapshot, which includes its changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS]. Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?  wiresong.ca}

Starting up \openSmalltalkVM from source files has never been supported. However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files. The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.

\paragraph{Interpreter-only simulation.} 
As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT. Since executing machine code on the processor simulations is not fast, interpreter simulation may be faster, and startup time and memory footprint are smaller.
